---
title: "MMVshiny: a step-by-step demo app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MMVshiny: a step-by-step demo app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

This vignette is a guided, *step-by-step* walkthrough of the **MMVshiny demo app** that
ships with the package.

The demo mirrors the architecture used in the larger MMV applications (MMVSola/MMVFree):

- inputs are specified in a **state specification** (an Excel sheet)
- MMVshiny generates JavaScript handlers and R scripts for observers/reactives
- the server processes GUI events via a single **state** object

## Get the demo app files

The demo app is distributed as a zip file in `inst/extdata/`.

```{r locate-demo}
zip_path <- system.file("extdata", "MMVshinyDemo.zip", package = "MMVshiny")
stopifnot(nzchar(zip_path))
zip_path
```

Unzip it to a temporary folder so we can inspect the files.

```{r unzip-demo}
app_dir <- file.path(tempdir(), "MMVshinyDemo")
if (dir.exists(app_dir)) unlink(app_dir, recursive = TRUE)
utils::unzip(zip_path, exdir = app_dir)
list.files(app_dir, recursive = TRUE)
```

You should see the following key files:

- `app.R` – the Shiny application
- `Resources/MMVshinyDemoParameters.xlsx` – the **state specification**
- `Resources/FUNCTION_RunSimulation.R` – a small helper implementing a toy simulation

## Step 1: Understand the state specification

The state specification is an Excel sheet that defines *what inputs exist* and *how MMVshiny should
handle them*.

In the demo, it lives here:

```{r state-spec-path}
spec_file <- file.path(app_dir, "MMVshinyDemo", "Resources", "MMVshinyDemoParameters.xlsx")
file.exists(spec_file)
```

Load the specification using the MMVshiny helper (this is the same entry point used by MMVSola/MMVFree).

```{r load-spec}
stateSpec <- MMVshiny::LoadStateSpecification(filename = spec_file)
stateSpec[, .(ID, TYPE, GUILABEL, UNIT, MIN, MAX, OUTOFBOUNDS, VALEXPR, RADIOVALUES, STATUSICON, ACTIONEXPR)]
```

### What do the columns mean?

The demo uses only a subset of features, but the full sheet follows a consistent schema.
Below is a practical description of the columns you will most commonly use.

#### Identity and labels

- **`ID`**: the unique identifier for the parameter. This is also used as the GUI id by default.
- **`GUILABEL`**: label shown next to the input.
- **`REPORTLABEL`**: label used in reports (tables, exports).

#### Input type and UI choices

- **`TYPE`**: determines how MMVshiny creates the UI and how events are handled.
  Common values include `numeric input`, `text input`, `select input`, `radio input`, `action button input`.
- **`RADIOVALUES`**: options for radio/select inputs. In the demo it uses the format
  `Once daily:Twice daily`.

#### Units and formatting

- **`UNIT`**: a unit string (displayed in UI and/or reports).
- **`DECDIGITS`**: number of decimals to display.

#### Validation

- **`MIN` / `MAX`**: numeric bounds.
- **`MINNOTE` / `MAXNOTE`**: free-text explanations of the bounds.
- **`OUTOFBOUNDS`**: what to do if the user value violates bounds.
  For example, `keep with error` keeps the value but marks the input as invalid.

#### Defaults, reactive defaults, and provenance

- **`VALEXPR`**: an expression that determines the *default* (or a *reactive default*) for an input.
  In this demo, `Regimen` has a **reactive** default that depends on `Weight_kg`.
- **`DEFAULTVALUENOTE`**: explanation for the default.
- **`SOURCE`**: where the value came from (e.g. user input vs. generated/default).

#### Status icons and writeability

- **`STATUSICON`**: whether MMVshiny should render a small status icon next to the input (valid/invalid/source).
- **`WRITEABLE`**: whether the user can edit the input.

#### Scenario calculation fields

These columns are used in some MMV apps to support *scenario calculations*:

- **`SCFILTER`**, **`SCVALUE`**

They are present for schema consistency; the demo does not use them.

#### Action buttons

- **`ACTIONEXPR`**: only relevant for `action button input` rows. This is an R function that MMVshiny can call
  when the button event is processed.

In the demo, the `RunSim` row defines an action that runs a toy simulation and stores results into the state.

## Step 2: Walk through `app.R`

Open the app source (we show it as text here, but you can open it in any editor):

```{r app-source-path}
app_file <- file.path(app_dir, "MMVshinyDemo", "app.R")
file.exists(app_file)
```

Below we follow the same section headers used in `app.R` (comments ending in `----`).

### LOAD INPUT SPEC ----

```{r load-input-spec, eval=FALSE}
stateSpec <- MMVshiny::LoadStateSpecification(
  filename = "Resources/MMVshinyDemoParameters.xlsx"
)
```

This is the single source of truth for:

- which inputs exist
- how they should be validated
- how defaults are computed
- (optionally) what action buttons do

### GENERATE SCRIPTS ----

MMVshiny generates the “plumbing” code at app start.

#### Generate reactive getters inside the state ----

This creates functions like `state$Weight_kg()` and `state$Compound()` so your server logic
can read inputs in a consistent way.

```{r gen-reactives, eval=FALSE}
scriptCreateReactivesInState.R <- MMVshiny::GenerateScriptCreatingReactives(
  stateSpec,
  stateObjectName = "MYSTATE",
  filename = "Resources/Generated/CreateReactivesInState.R"
)
```

#### Generate observers inside the state ----

The demo generates two kinds:

1. **Reset observers**: initialize (and re-initialize) all values by processing a reset event.
2. **Default observers**: evaluate reactive defaults (here only `Regimen`).

```{r gen-state-observers, eval=FALSE}
scriptObserveResetCountInState.R <- MMVshiny::GenerateScriptCreatingObservers(
  stateSpec,
  stateObjectName = "MYSTATE",
  observerTypes = c("resetCount"),
  filename = "Resources/Generated/ObserveResetCountInState.R"
)

scriptObserveDefaultInState.R <- MMVshiny::GenerateScriptCreatingObservers(
  stateSpec,
  stateObjectName = "MYSTATE",
  observerTypes = c("default"),
  ids = c("Regimen"),
  filename = "Resources/Generated/ObserveDefaultInState.R"
)

scriptsToSourceInState <- c(
  scriptCreateReactivesInState.R,
  scriptObserveResetCountInState.R,
  scriptObserveDefaultInState.R
)
```

#### Generate observers in the server ----

These observers do two key jobs:

- respond to user edits (via `count<ID>` inputs) and call `ProcessGuiInputEvent()`
- push state changes back to the GUI (the `displayed` observer)

```{r gen-server-observers, eval=FALSE}
scriptObserveEventsInServer.R <- MMVshiny::GenerateScriptCreatingObservers(
  stateSpec,
  stateObjectName = "state",
  observerTypes = c("countID", "displayed"),
  filename = "Resources/Generated/ObserveEventsInServer.R"
)
```

#### Generate status icons ----

If `STATUSICON == TRUE`, MMVshiny can render small icons next to inputs to communicate
validity and source.

```{r gen-status-icons, eval=FALSE}
scriptRenderStatusIcons.R <- MMVshiny::GenerateScriptRenderingIcons(
  stateSpec,
  prefixGuiId = "",
  stateObjectName = "state",
  filename = "Resources/Generated/RenderStatusIcons.R"
)
```

### UI ----

MMVshiny relies on a small piece of JavaScript to detect when a user has “committed” a change
(press Enter or click outside a text field). The handler increments `input$count<ID>`.

```{r gen-js, eval=FALSE}
scriptHandleEvents.js <- MMVshiny::GenerateJavaScriptEventHandlers(
  stateSpec,
  filename = "www/Generated/HandleEvents.js"
)

ui <- fluidPage(
  includeScript(scriptHandleEvents.js),
  ...
)
```

The actual inputs are created from the spec using `CreateUIInput()`.

```{r create-ui-inputs, eval=FALSE}
tags$table(
  tags$tr(MMVshiny::CreateUIInput(spec = stateSpec, id = "Compound", useLabel = TRUE)),
  tags$tr(MMVshiny::CreateUIInput(spec = stateSpec, id = "Weight_kg", useLabel = TRUE)),
  tags$tr(MMVshiny::CreateUIInput(spec = stateSpec, id = "Height_cm", useLabel = TRUE)),
  tags$tr(MMVshiny::CreateUIInput(spec = stateSpec, id = "Regimen", useLabel = TRUE))
)

MMVshiny::CreateUIInput(spec = stateSpec, id = "RunSim")
```

### SERVER ----

The server starts by creating a single state object.

```{r init-state, eval=FALSE}
state <- MMVshiny::InitState(
  stateSpec,
  listObjects = list(
    prefixGuiId = "",
    result = shiny::reactiveVal(NULL),
    lastSummary = shiny::reactiveVal(NULL)
  ),
  scriptsToSource = scriptsToSourceInState
)
```

Then it sources the generated scripts:

```{r source-generated, eval=FALSE}
source(scriptRenderStatusIcons.R, local = TRUE)
source(scriptObserveEventsInServer.R, local = TRUE)
```

#### Where does the simulation run?

The state spec row `RunSim` is of type `action button input`.
When the user clicks the button, MMVshiny processes the GUI event and (because `ACTIONEXPR` is defined)
executes the action function.

That action stores results into:

- `state$result()` – time-course data for plotting / tables
- `state$lastSummary()` – a small summary list used by `output$summary`

Your server outputs then become straightforward:

```{r outputs, eval=FALSE}
output$summary <- renderText({
  s <- state$lastSummary()
  if (is.null(s)) return("Click 'Run simulation' to generate results.")
  paste0(
    "Compound: ", state$Compound(), "\n",
    "Regimen: ", s$regimen, "\n",
    "BMI: ", round(s$bmi, 1), "\n",
    "Suggested daily dose (toy): ", s$dose_mg, " mg"
  )
})

output$plot <- renderPlot({
  res <- state$result()
  if (is.null(res) || nrow(res) == 0) return(NULL)
  plot(res$time_h, res$conc, type = "l", xlab = "Time (h)", ylab = "Concentration (a.u.)")
})
```

## Step 3: Run the demo app

To run the demo from an installed MMVshiny package:

```{r run-demo, eval=FALSE}
zip_path <- system.file("extdata", "MMVshinyDemo.zip", package = "MMVshiny")
app_dir <- file.path(tempdir(), "MMVshinyDemo")
if (dir.exists(app_dir)) unlink(app_dir, recursive = TRUE)
utils::unzip(zip_path, exdir = app_dir)

# The zip contains the folder "MMVshinyDemo" which includes app.R
shiny::runApp(file.path(app_dir, "MMVshinyDemo"))
```

## Where to go next

To adapt this pattern for your own app:

1. Create your own Excel state specification.
2. Use `CreateUIInput()` to generate the UI elements.
3. Generate scripts for observers/reactives and source them.
4. Put your domain logic in:
   - reactive defaults (`VALEXPR`)
   - action buttons (`ACTIONEXPR`)
   - server outputs consuming `state$<ID>()` getters.
